---
title: "Using the gridsearch Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the gridsearch Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = 'center',
  dpi = 500,
  out.width = "90%"
)

library(dplyr)
library(gridsearch)
library(ggplot2)

```

# Motivation

Researchers working with panel data on public opinion generally want to know the answers to two basic questions: "how many people changed?" and "how much they changed?"

That said, longitudinal change could result from various mechanisms: large segments of the population may shift their opinions in a specific direction in small or large amounts; a small segment of the population may have large changes; or a mix of changes---including positive *and* negative movements---may alter the overall balance.

The `gridsearch` is designed to give researchers a preliminary check as to what potential data generation processes may have generated their observed data.

# Illustration

## Data Generation Processes

Think about a panel study, with an underlying data-generation process (DGP). This panel study consists of 1,000 individuals observed across 3 time periods. Since this is a simulated data, we know that 50% of these 1,000 individuals *changed* their underlying latent positions within this window with 1 SD in the positive direction. However, there are reliability issues with this specific item (let's say it's measured with a reliability of 90%) and the response resolution is 2 (suppose people only say "agree" or "disagree").

```{r}
# load the data
d <- readRDS("./simulated_data.rds")

# glimpse
head(d)
```

As we see, there are individuals (`pid`) observed on `y` across `t`.

Over a three-wave period, we have multiple response patterns observed in our empirical data or, more specifically, we have 8 potential trajectories people can walk through):

```{r}
# load the patterns
p <- readRDS("./simulated_patterns.rds")

# glimpse
print(p)
```

Given this setup, how can we answer the two questions we initially posed?

## Grid Search Algorithm

The grid-search algorithm tries to answer this question. It performs six main steps:

1) Simulates data with several parameters of interest for our DGPs---rate of change, strength of change, direction of change, and reliability of response measurement using

   $$
   y_{it} = Y_i + \delta \tau_i D_{it} + \epsilon_{it} \ \text{ with } \epsilon_{it} \sim \mathcal{N}(0, 1) \tag{DGP}
   $$

2) Calculates the distribution of response trajectories in both the observed and simulated data, by either tabulating response patterns across time or estimating slope coefficients;

3) Compares the simulated trajectories to the observed trajectories. It does so by calculating a "distance" score to quantify the extent to which specific simulated values reproduce the observed values;

4) Rejects the simulated data if the distance measure is above a certain pre-specified "caliper" value, and accepts it otherwise;

5) Replicates this analysis $N$ times by drawing samples from the specified priors; 

6) Inspects the parameter combinations that most closely approximate to the observed values.

Let's apply this to our simulated dataset.

```{r}
grid <- gridSearch(
  data = d,
  yname = 'y', ## observed binary outcome
  tname = 'time', ## time variable
  pname = 'pid', ## panel identifier
  n_samples = 1e4, ## the number of draws from the priors
  caliper = .15 ## the distance to accept or reject specific samples
)

head(grid)
```

The `gridSearch` function gives us a data table with four columns:

- `ic_sample` shows the value of the `strength` parameter informing our DGP,
- `pc_sample` shows the value of the `rate` parameter informing our DGP,
- `balance` shows the extent to which our sample makes positive changes; put differently, it shows the `direction` of change in the sample,
- `rel_sample` provides a value of the reliability score retrieved from the search.

Each row is one DGP that is "accepted" by our protocol with a given caliper value.

## Evaluating the Potential DGPs

We can visually inspect the parameter distributions using `gridPlot`:

```{r}
gridPlot(grid, type = "posterior")
```

Alternatively, we can look at the conditional accepted samples:

```{r}
gridPlot(grid, type = "conditional")
```

# Warnings and Notes

We would like to flag a few issues that you need to be aware of:

- Currently, `gridsearch` only accommodates binary outcomes. We are hoping to generalize these procedures to cases where we have more than 2 response categories.
- While the `slopes` method in the `gridSearch` function may be helpful in longer panels, we note that its properties are less worked out than the default method.

For more details about the protocol, please see the article, *The Promises and Pitfalls of Using Panel Data to Understand Individual Belief Change*, stored in [SocArXiv](https://osf.io/preprints/socarxiv/rhf4q).
